---
title: "Calculate Reliability"
author: "Chandresh Lokesha"
date: "Spring 24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

An analytics group has built a distributed data lake as part of a new data warehouse. The data warehouse consists of a number of distinct components, including:

- A database on a cloud server (99.987% availability)
- Two identical database management system servers (99.3%)
- An analytics logic server (94.7%)
- Network infrastructure to the cloud server (98.37%)

What is the overall availability of the system, expressed as a percentage with three digits of precision? Show your work.

### Solution:

The overall availability of the system can be calculated using the formula for parallel systems:

\[
A_{\text{overall}} =\prod_{i=1}^{n} (A_i)
\]

where \( A_i \) represents the availability of each component.

Given:

- Cloud server availability (\( A_1 \)) = 99.987%
- Database management system servers availability (\( A_2 \)) = 99.3%
- Analytics logic server availability (\( A_3 \)) = 94.7%
- Network infrastructure availability (\( A_4 \)) = 98.37%

We can calculate the overall availability:

```{r}
cloud_server <- 0.99987
db_servers <- 0.993
analytics_server <- 0.947
network_infrastructure <- 0.9837

overall_availability <- cloud_server * db_servers * analytics_server * network_infrastructure
overall_availability_percentage <- overall_availability * 100
round(overall_availability_percentage, 3)
```


## Question 2
### Solution:

The expected annual reliability of the RAID 1+0 storage system can be calculated using the formula:

\[
\text{Reliability} = e^{-\frac{\text{hours}}{\text{MBTF}}}
\]

Given:

- MBTF = 1,450,000 hours

We can calculate the expected annual reliability:

```{r}
MBTF <- 1450000
hours_per_year <- 24 * 365

reliability <- exp(-hours_per_year / MBTF)
reliability

```
