---
title: "Assignment / Explore Query Planning and Indexing"
author: "Chandresh Lokesha"
date: "Spring 2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r connectToDB, warning=FALSE}
library(RSQLite)

# Step 1: Connect to the SQLite database
conn <- RSQLite::dbConnect(RSQLite::SQLite(), dbname = "sakila.db")

```

```{r testQuery, eval=FALSE}
res <- dbGetQuery(conn, "SELECT * from FILM")
print(res)
```

## Question 1
```{r Q1, warning=FALSE}
# Step 2: Retrieve the list of user-defined indexes
indexes <- RSQLite::dbGetQuery(conn, 
                               "SELECT name FROM sqlite_master 
                               WHERE type = 'index' 
                               AND tbl_name != 'sqlite_sequence' 
                               AND sql NOT LIKE '%UNIQUE%' 
                               AND sql NOT LIKE '%PRIMARY KEY%'")

# Step 3: Drop each user-defined index
if (nrow(indexes) > 0) {
  for (i in 1:nrow(indexes)) {
    RSQLite::dbExecute(conn, paste0("DROP INDEX IF EXISTS ", indexes$name[i]))
  }
}

res <- dbGetQuery(conn, "SELECT l.name, count(*) from FILM f, LANGUAGE l 
                  where l.language_id = f.language_id GROUP BY l.language_id")

print(res)
```
## Question 2
```{r Q2_QueryPlan, warning=FALSE}
query <- "SELECT l.name, count(*) AS film_count 
          FROM FILM f 
          JOIN LANGUAGE l ON l.language_id = f.language_id 
          GROUP BY l.name"
plan <- RSQLite::dbGetQuery(conn, paste("EXPLAIN QUERY PLAN", query))
print(plan)
```

## Question 3
```{r Q3_ZorroArk, warning=FALSE}
query3 <- "SELECT f.title, f.length, c.Name AS film_category
          FROM FILM f 
          JOIN FILM_CATEGORY fc on fc.film_Id = f.film_Id
          JOIN Category c on fc.category_id = c.category_id
          where f.title = 'ZORRO ARK'"

res <- RSQLite::dbGetQuery(conn, query)
print(res)
```

## Question 4
```{r Q4_Q3QueryPlan}

plan <- RSQLite::dbGetQuery(conn, paste("EXPLAIN QUERY PLAN", query3))
print(plan)
```

## Question 5
```{r, createIndexTitle}
RSQLite::dbExecute(conn, "CREATE INDEX IF NOT EXISTS TitleIndex ON FILM (TITLE)")

```

## Question 6
```{r, indexedQuery3}
plan <- RSQLite::dbGetQuery(conn, paste("EXPLAIN QUERY PLAN", query3))
print(plan)
```
## Question 7 Comments and differences
### Yes, they are different.Looking at the query plan on the film table details we can say that indxing is being used in the later. As you can see before creating index SEARCH on f is performed using primary key, and after creating index search f is based on TitleIndex

## Question 8
## Measure Time without Index Title
```{r measureTimeWithoutIndex, warning=FALSE}
dbExecute(conn, "DROP INDEX IF EXISTS TitleIndex;")
bt <- Sys.time()
execQuery <- function ()
{
 res <- RSQLite::dbSendQuery(conn, query3)
}

et <- Sys.time()

t.which <- et - bt

cat("Time elapsed: ", round((t.which),3), " sec")

```

## Measure Time with Index title
```{r Measuretime, warning=FALSE}
RSQLite::dbExecute(conn, "CREATE INDEX IF NOT EXISTS TitleIndex ON FILM (TITLE)")
bt <- Sys.time()
execQuery <- function ()
{
 res <- RSQLite::dbSendQuery(conn, query3)
}

et <- Sys.time()

t.which <- et - bt

cat("Time elapsed: ", round((t.which),3), " sec")


```



## Report: time taken for the entire expression to execute
### As you can see the indexing has reduced time to execute

## Question 9
```{r Q9usingLikeOp, warning=FALSE}

query1 <- "SELECT l.name, f.title, f.length 
            from FILM f, LANGUAGE l 
            where l.language_id = f.language_id and title LIKE '%[Gg][Oo][Ll][Dd]%';"
            

res <- dbGetQuery(conn, query1)

```

## Question 10
```{r Q10_q9QueryPlan, warning=FALSE}
plan <- RSQLite::dbGetQuery(conn, paste("EXPLAIN QUERY PLAN", query1))
print(plan)

```
### There is no difference i.e it did not use the index (as you can see in in the query plan deatils 'SCAN f' is performed).This is because in the query plan 'LIKE' operator is used and it will not make use of indexing