---
title: "Analysis of Wildlife Strikes to Aircraft"
author: "Chandresh Lokesha"
date: "Spring 24"
output:
  pdf_document: default
  html_document: default
---

```{r disconnectAllLocalConn, eval = F, warning=FALSE}
library(RMySQL)  

killDbConnections <- function () {

  all_cons <- dbListConnections(MySQL())

  print(all_cons)

  for(con in all_cons)
    +  dbDisconnect(con)

  print(paste(length(all_cons), " connections killed."))

}
killDbConnections()
```

## Practicum I CS5200

## Connect to local DB
```{r connectToDB, eval = T, warning = F}
# Clearing the global vars
rm(list = ls())

# imports and connecting to DB
library(DBI)
library(RMySQL)
library(lubridate)
library(dplyr)
birdStrikeDBCon = dbConnect(RMySQL::MySQL(),
                            dbname='birdstrike',
                            host='localhost',
                            port=3306,
                            user='root',
                            password='password')

```

## Connect to online DB
```{r connectToOnlineDB, eval = F, warning = F}
# Clearing the global vars
rm(list = ls())

# imports and connecting to DB
library(DBI)
library(RMySQL)
library(lubridate)
library(dplyr)
birdStrikeDBCon = dbConnect(RMySQL::MySQL(),
                            dbname='sql5694564',
                            host='sql5.freemysqlhosting.net',
                            port=3306,
                            user='sql5694564',
                            password='wd6xs7ESMR')

```

### not required for connecting to online DB
```{sql connection=birdStrikeDBCon, eval = F, warning = F}
# to enable batch uploading of CSV to DB from client side
set global local_infile = TRUE;
```

```{r DropTablesBeforeCreation, eval = T, warning = F}

# Drop the tables if they already exist
dbSendQuery(birdStrikeDBCon, "DROP TABLE IF EXISTS strikes;")
dbSendQuery(birdStrikeDBCon, "DROP TABLE IF EXISTS conditions;")
dbSendQuery(birdStrikeDBCon, "DROP TABLE IF EXISTS flights;")
dbSendQuery(birdStrikeDBCon, "DROP TABLE IF EXISTS airports;")
```

## Q4 creating tables 
```{r CreateTables, eval = T, warning = F}

# Create table airport
dbSendQuery(birdStrikeDBCon, "CREATE TABLE airports (
  aid INTEGER PRIMARY KEY AUTO_INCREMENT,
  airportName TEXT,
  airportState TEXT,
  airportCode VARCHAR(3) DEFAULT 'ZZZ'
);
")

# Create table Flights
dbSendQuery(birdStrikeDBCon, "CREATE TABLE flights (
  fid INTEGER PRIMARY KEY AUTO_INCREMENT,
  date DATE,
  originAirport INTEGER,
  airlineName TEXT,
  aircraftType TEXT,
  isHeavy BOOLEAN,
  FOREIGN KEY (originAirport) REFERENCES airports(aid)
);
")

# Create table conditions
dbSendQuery(birdStrikeDBCon, "CREATE TABLE conditions (
  cid INTEGER PRIMARY KEY AUTO_INCREMENT,
  sky_condition TEXT,
  explanation TEXT
);
")

# create table strikes
dbSendQuery(birdStrikeDBCon, "CREATE TABLE strikes (
  sid INTEGER PRIMARY KEY AUTO_INCREMENT,
  fid INTEGER,
  numbirds INTEGER,
  impact TEXT,
  damage BOOLEAN,
  altitude INTEGER CHECK (altitude >= 0),
  conditions INTEGER,
  FOREIGN KEY (fid) REFERENCES flights(fid),
  FOREIGN KEY (conditions) REFERENCES conditions(cid)
);
")
```

```{r testTables, eval=FALSE}

# Insert sample data into the 'airports' table
dbSendQuery(birdStrikeDBCon, "INSERT INTO airports (aid, airportName, airportState, airportCode) VALUES (1, 'Airport1', 'State1', 'ABC');")
res <- dbExecute(birdStrikeDBCon, "SELECT * from airports")
print(res)

# Insert sample data into the 'flights' table
dbSendQuery(birdStrikeDBCon, "INSERT INTO flights (fid, date, originAirport, airlineName, aircraftType, isHeavy) VALUES (2, '2024-03-06', 1, 'Airline1', 'Aircraft1', TRUE);")
res <- dbExecute(birdStrikeDBCon, "SELECT * from flights")
print(res)

# Insert sample data into the 'conditions' table
dbSendQuery(birdStrikeDBCon, "INSERT INTO conditions (cid, sky_condition, explanation) VALUES (2, 'Overcast', 'Explanation1');")
res <- dbExecute(birdStrikeDBCon, "SELECT * from conditions")
print(res)

# Insert sample data into the 'strikes' table
dbSendQuery(birdStrikeDBCon, "INSERT INTO strikes (sid, fid, numbirds, impact, damage, altitude, conditions)
VALUES (3, 2, 5, 'Impact1', TRUE, 10000, 2);")
res <- dbExecute(birdStrikeDBCon, "SELECT * from strikes")
print(res)

# Fetch and print data from the 'airports' table
airports_data <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM airports;")
print(airports_data)

# Fetch and print data from the 'flights' table
flights_data <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM flights;")
print(flights_data)

# Fetch and print data from the 'conditions' table
conditions_data <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM conditions;")
print(conditions_data)

strikes_data <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM strikes;")
print(strikes_data)

dbSendQuery(birdStrikeDBCon, "DELETE FROM strikes;")
dbSendQuery(birdStrikeDBCon, "DELETE FROM flights;")
dbSendQuery(birdStrikeDBCon, "DELETE FROM airports;")
q4 <- dbSendQuery(birdStrikeDBCon, "DELETE FROM conditions;")
dbClearResult(q4)
```

## Q5 load CSV into dataframe
```{r LoadCSV, eval = T, warning = F}
# Load CSV file into a dataframe
bds.raw <- read.csv("BirdStrikesData-V3.csv", header = TRUE, sep = ",")
```

### drop half of the rows as the db size exceeds 5mb on mysqlfree.net and suspends the account
```{r dropHalfTofitOnlineDB, eval=F, warning=FALSE}
bds.raw <- bds.raw[(nrow(bds.raw)/2 + 1):nrow(bds.raw), ]
```

```{r printCSV, eval=FALSE}
print(head(bds.raw))
print(colnames(bds.raw))
print(nrow(bds.raw))
print(nrow(unique(bds.raw)))
```

## Data Cleaning and pre-processing to format the date

### 1. We are normalizing the airline, airport using 'unknown'
```{r parseDateforDeparture, eval = T, warning = F}
bds.raw$flight_date<- parse_date_time(bds.raw$flight_date, orders=c("m-d-y H:M","m/d/y H:M"))
bds.raw$flight_date <- format(bds.raw$flight_date,"%Y-%m-%d")
```

```{r cleaningSentinelData, eval = T, warning = F}
arilineCase <- which(bds.raw$airline=='')
bds.raw[arilineCase,"airline"] <- 'unknown'

aircraftCase <- which(bds.raw$aircraft=='')
bds.raw[aircraftCase,"aircraft"] <- 'unknown'

airportCase <- which(bds.raw$airport=='')
bds.raw[airportCase,"airport"] <- 'unknown'
```

### Add size mappings for wildlife size
```{r sizeMaping, eval = T, warning = F}
size_mapping <- c("Small" = 1, "Medium" = 2, "Large" = 3)
bds.raw$wildlife_size_numeric <- size_mapping[bds.raw$wildlife_size]
```

### omit flights without flight information
```{r omitEmptyFlightDetails, eval = T, warning = F}
bds.raw <- bds.raw[which(bds.raw$airport!='' & bds.raw$model!='' & bds.raw$origin!='' ),]
```

```{r testUpdatedData, eval = FALSE}
print(head(bds.raw))
print(colnames(bds.raw))
print(nrow(bds.raw))
```
## Insert the data into airports table
```{r insertIntoAirports, eval = T, warning = F}
# Function: insert_airports
# Description: This function inserts airport data into the 'airports' table.
#
# Parameters:
#   df - Dataframe containing airport information.
#   dbconn - database connection string
insert_airports <- function(dbconn, df) {
  # Transform and clean data if needed
  
  df <- df[, c("airport", "origin")]
  
  df$airport <- trimws(df$airport)
  
  colnames(df) <- c("airportName", "airportState")
  
  # Create a unique set of airports
  unique_airports <- unique(df)
  
  # Add synthetic primary key 'aid'
  # unique_airports$aid <- seq_len(nrow(unique_airports))
  
  # Insert data into the 'airports' table
  dbWriteTable(dbconn, "airports", unique_airports, append = TRUE, row.names = FALSE)
}

# Insert data into the 'airports' table
insert_airports(birdStrikeDBCon, bds.raw)

```

```{r testAirportTableInsert, eval = FALSE}
# Display part of the 'airports' table
partial_airports <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM airports LIMIT 5")
print(partial_airports)
total_rows <- dbGetQuery(birdStrikeDBCon, "SELECT COUNT(*) FROM airports")
print(total_rows)
partial_airports <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM airports where airportName = 'ABERDEEN REGIONAL AR'")
print(partial_airports)
```

## Insert the data into flights table
```{r insertIntoFlights, eval = T, warning = F}
# Function: insert_flights
# Description: Inserts flight data into the 'flights' table.
#
# Parameters:
#   conn - Database connection.
#   df - Dataframe containing flight information.
insert_flights <- function(conn, df) {
  # Transform and clean data if needed
  # print(df)
  df <- df[, c("airline", "flight_date", "origin",  "aircraft", "heavy_flag")]
  colnames(df) <- c("airlineName", "date", "origin",  "aircraftType", "isHeavy")
  #print(df)
  df$date <- as.Date(df$date)  # Convert date to Date type
  
  # Map 'isHeavy' values to boolean
  df$isHeavy <- tolower(df$isHeavy) == "yes"
  
  # Map 'originAirport' values to corresponding 'aid' values from the 'airports' table
  airports_mapping <- dbGetQuery(conn, "SELECT * FROM airports")
  df$originAirport <- airports_mapping$aid[match(df$origin, airports_mapping$airportState)]
  
  df$origin <- NULL
  
  # Create a unique set of flights
  unique_flights <- unique(df)
  
  # Add synthetic primary key 'fid'
  # unique_flights$fid <- seq_len(nrow(unique_flights))
  
  # Insert data into the 'flights' table
  dbWriteTable(conn, "flights", unique_flights, append = TRUE, row.names = FALSE)
}

# insert into flights
insert_flights(birdStrikeDBCon, bds.raw)
```

```{r testFlightsTable, eval=FALSE}
# Display part of the 'flights' table
partial_airports <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM flights LIMIT 5")
print(partial_airports)

total_cnt_airlines <- dbGetQuery(birdStrikeDBCon, "SELECT COUNT(*) FROM flights")
print(total_cnt_airlines)
```


### Inserting into conditions table
```{r insertIntoConditions, eval = T, warning = F}
# Function: insert_conditions
# Description: Inserts sky conditions data into the 'conditions' table.
#
# Parameters:
#   conn - Database connection.
#   df - Dataframe containing sky conditions information.

insert_conditions <- function(conn, df) {
  # Transform and clean data if needed
  df <- df[, c("sky_conditions")]
  
  # Create a data frame with the desired column name
  unique_conditions <- unique(data.frame(sky_condition = df))
  
  # Add an empty 'explanation' column
  unique_conditions$explanation <- NA
  
  # Add synthetic primary key 'cid'
  # unique_conditions$cid <- seq_len(nrow(unique_conditions))
  
  # print(unique_conditions)
  
  # Insert data into the 'conditions' table
  dbWriteTable(conn, "conditions", unique_conditions, append = TRUE, row.names = FALSE)
}

insert_conditions(birdStrikeDBCon, bds.raw)
```

```{r testConditionsTable, eval = FALSE}
# Display part of the 'conditions' table
partial_conditions <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM conditions LIMIT 5")
print(partial_conditions)

total_cnt_condts <- dbGetQuery(birdStrikeDBCon, "SELECT COUNT(*) FROM conditions")
print(total_cnt_condts)
```

### inserting into strikes table
```{r insertIntoStrikes, eval = T, warning = F}
# Function: insert_strikes
# Description: Inserts bird strike data into the 'strikes' table.
#
# Parameters:
#   conn - Database connection.
#   df - Dataframe containing bird strike information.

insert_strikes <- function(conn, df) {
  df <- df[, c("flight_date","wildlife_size_numeric", "impact", "damage", "altitude_ft", "sky_conditions", "aircraft", "airline", "heavy_flag", "origin")]
  colnames(df) <- c("date","numbirds", "impact", "damage", "altitude", "conditions", "aircraft", "airline", "isHeavy", "origin")
  
  df$isHeavy <- tolower(df$isHeavy) == "yes"
  
  # Map 'damage' values to boolean
  df$damage <- tolower(df$damage) == "damage"

  
  # Convert altitude to integer
  df$altitude <- as.integer(df$altitude)
  
  # Map 'airport' to corresponding 'fid' values from the 'flights' table
  flights_mapping <- dbGetQuery(conn, "SELECT * from flights")
  
  df$fid <- flights_mapping$fid[match(
  df$date, flights_mapping$date) & 
  match(df$aircraft, flights_mapping$aircraftType) & 
  match(df$airline, flights_mapping$airlineName) &
  match(df$isHeavy, flights_mapping$isHeavy)]
  
  # Map 'sky_conditions' to corresponding 'cid' values from the 'conditions' table
  conditions_mapping <- dbGetQuery(conn, "SELECT * FROM conditions")
  df$conditions <- conditions_mapping$cid[match(df$conditions, conditions_mapping$sky_condition)]
  
  # Create a unique set of strikes
  unique_strikes <- (df)
  # print(length(unique_conditions))  # Print the count of unique values
  # Add synthetic primary key 'sid'
  # unique_strikes$sid <- seq_len(nrow(unique_strikes))
  
  # Remove 'date', 'aircraft', and 'airlineName' columns before inserting into the 'strikes' table
  columns_to_remove <- c("date", "aircraft", "airline", "isHeavy", "origin")
  unique_strikes <- unique_strikes[, !(names(unique_strikes) %in% columns_to_remove)]

  # Insert data into the 'strikes' table
  dbWriteTable(conn, "strikes", unique_strikes, append = TRUE, row.names = FALSE)
}

insert_strikes(birdStrikeDBCon, bds.raw)
```

```{r testStrikesTable, eval=FALSE}
# Display part of the 'strikes' table
partial_strikes <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM strikes LIMIT 5")
print(partial_strikes)
total_cnt_strikes <- dbGetQuery(birdStrikeDBCon, "SELECT COUNT(*) FROM strikes")
print(total_cnt_strikes)
```

## Q8 Top Airports with Strikes

```{sql, connection = birdStrikeDBCon}

SELECT a.airportState, COUNT(s.sid) as incidentCount
FROM airports a
JOIN flights f ON a.aid = f.originAirport
JOIN strikes s ON f.fid = s.fid
GROUP BY a.airportState
ORDER BY incidentCount DESC
LIMIT 10;
```

## Analysis by Airline
### Online DB on freeSql.net is not supporting WITH i.e CTE clause and hence added another chunk below
```{sql, eval = F, connection = birdStrikeDBCon}

WITH AirlineIncidentCounts AS (
  SELECT f.airlineName, COUNT(s.sid) as incidentCount
  FROM flights f
  JOIN strikes s ON f.fid = s.fid
  GROUP BY f.airlineName
)

SELECT airlineName, incidentCount
FROM AirlineIncidentCounts
WHERE incidentCount > (SELECT AVG(incidentCount) FROM AirlineIncidentCounts)
ORDER BY incidentCount DESC;
```
```{sql, connection = birdStrikeDBCon}
SELECT f.airlineName, COUNT(s.sid) AS incidentCount
FROM flights f
JOIN strikes s ON f.fid = s.fid
GROUP BY f.airlineName
HAVING COUNT(s.sid) > (
    SELECT AVG(incidentCount)
    FROM (
        SELECT COUNT(sid) AS incidentCount
        FROM strikes
        GROUP BY fid
    ) AS subquery
)
ORDER BY incidentCount DESC;
```

## Analysis by Month on numberOfStrikes

```{r monthlyTrend, eval = T, warning = F}
# Execute SQL query and store the result in a dataframe
monthly_analysis_by_strikes <- dbGetQuery(birdStrikeDBCon, "
  SELECT
    MONTHNAME(f.date) AS Month,
    COUNT(s.sid) AS NumberOfStrikes
  FROM
    flights f
  JOIN
    strikes s ON f.fid = s.fid
  GROUP BY
    MONTHNAME(f.date)
  ORDER BY
    NumberOfStrikes DESC;
")
```

```{r testMonthlyTrend, eval=FALSE}
# Display the dataframe
print(monthly_analysis_by_strikes)
```

## Analysis by NumberOfBirds
```{r monthlyNumberOfBirds, eval = T, warning = F}
monthly_analysis <- dbGetQuery(birdStrikeDBCon, "SELECT
  MONTHNAME(f.date) AS Month,
  SUM(s.numbirds) AS NumberOfBirds
FROM
  flights f
JOIN
  strikes s ON f.fid = s.fid
GROUP BY
  MONTHNAME(f.date)
ORDER BY
  NumberOfBirds DESC;
")

# print(monthly_analysis)
```

## Trend by Month of Number of Birds
```{r visualizationByNumberofBirds, eval = T, warning = F}
library(ggplot2)

# Assuming 'monthly_analysis' is the dataframe from Question 10
# Define the order of months
month_order <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# Convert 'Month' to a factor with specified levels
monthly_analysis$Month <- factor(monthly_analysis$Month, levels = month_order)

# Order the dataframe by the new factor levels
monthly_analysis <- monthly_analysis[order(monthly_analysis$Month), ]

df <- monthly_analysis
f <- ggplot(df, aes(x = Month, y = NumberOfBirds))
f + geom_col(fill = "#0073C2FF") +
  geom_text(aes(label = NumberOfBirds), vjust = -0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Trend by Month of Number of Strikes
```{r visualizationByNoOfStrikes, eval = T, warning = F}
library(ggplot2)

monthly_analysis <- monthly_analysis_by_strikes
# Assuming 'monthly_analysis' is the dataframe from Question 10
# Define the order of months
month_order <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# Convert 'Month' to a factor with specified levels
monthly_analysis$Month <- factor(monthly_analysis$Month, levels = month_order)

# Order the dataframe by the new factor levels
monthly_analysis <- monthly_analysis[order(monthly_analysis$Month), ]

df <- monthly_analysis
f <- ggplot(df, aes(x = Month, y = NumberOfStrikes))
f + geom_col(fill = "#0073C2FF") +
  geom_text(aes(label = NumberOfStrikes), vjust = -0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{sql, connection=birdStrikeDBCon}
-- Create the audit_log table if it does not exist
CREATE TABLE IF NOT EXISTS audit_log (
    log_id INT AUTO_INCREMENT PRIMARY KEY,
    modification_type VARCHAR(50),
    table_name VARCHAR(50),
    timestamp TIMESTAMP DEFAULT CURRENT_TIMESTAMP,
    removed_sid INT
);
```
```{r}
# Drop the existing procedure if it exists

dbExecute(birdStrikeDBCon, "DROP PROCEDURE IF EXISTS RemoveStrikeAndLog;")
```

```{sql,connection=birdStrikeDBCon }

-- Create the stored procedure

CREATE PROCEDURE RemoveStrikeAndLog(IN sidToRemove INT)
BEGIN
    DECLARE removed_fid INT;
    
    -- Get the fid associated with the strike to be removed
    SELECT fid INTO removed_fid FROM strikes WHERE sid = sidToRemove;
    
    -- Log the removal in the audit_log table
    INSERT INTO audit_log (modification_type, table_name, removed_sid)
    VALUES ('Removal', 'strikes', sidToRemove);
    
    -- Remove the strike from the strikes table
    DELETE FROM strikes WHERE sid = sidToRemove;
    
    -- Optionally, additional cleanup/handle cascading deletes
    
    SELECT 'Strike removed successfully' AS result;
END 
```

```{r invokeProcedure, eval = T, warning = F}

# Call the stored procedure
dbSendQuery(birdStrikeDBCon, "CALL RemoveStrikeAndLog(1)")

```


```{r disconnect, eval = T, warning = F}
dbDisconnect(birdStrikeDBCon)
```

```{r testAuditLog,eval=FALSE}
# test for local
birdStrikeDBCon = dbConnect(RMySQL::MySQL(),
                            dbname='birdstrike',
                            host='localhost',
                            port=3306,
                            user='root',
                            password='password')
result <- dbGetQuery(birdStrikeDBCon, "SELECT * FROM audit_log")
print(result)
result2 <- dbGetQuery(birdStrikeDBCon, "SELECT * from strikes where sid = '1'")
print(result2)

# dbSendQuery(birdStrikeDBCon, "Delete from audit_log")
dbDisconnect(birdStrikeDBCon)
```